A tutorial on the model-dependent analysis of neutron and X-ray reflectometry

### 2020-06-22 refl_maths/paper.tex
mention jacobian
Update paper.tex
started some comments
test snakemake
remove mathcal

@@

When we measure a reflectometry profile, we measure the reflected intensity, and some uncertainty in that measurement, at discrete points in the wavevector, $R(q) \pm \delta R(q)$. 
Using the recursive method discussed above, we can calculate a model reflected intensity at these same $q$ values, $R_m(q)$. 
We then aim to reduce the difference between the measured and modelled reflected intensity through the optimisation (maximisation) of the likelihood, $\mathcal{L}$, 
%
\begin{equation}
    \mathcal{L} = \exp{\bigg\{-0.5 \sum_{q=q_{\text{min}}}^{q_{\text{max}}} \bigg[\frac{R(q) - R_m(q)}{\delta R(q)}\bigg]^2 + \ln[2\pi \delta R(q)]\bigg\}}.
\end{equation}
%
This parameter described the probability that the model $m$ accurately describes the observed reflectometry data. 
The aim in reflectometry modelling is to obtain a model reflectometry that has the maximum likelihood possible for the given experimental data. 
Figure~\ref{fig:likehood} shows the maximum likelihood model for some experimental data, in addition to another model which does not agree well with the data and therefore has a lower likelihood as a result. 
%
\begin{figure}[t]
    \includegraphics[width=0.49\textwidth]{likelihood}
    \caption{The maximum likelihood fit, the blue data points indicate some experimental data, while the yellow line shows the maximum likelihood model fit to this dataset and the green line show another model (which doesn't maximise the likelihood).}
    \label{fig:likelihood}
\end{figure}
%

@@

The final stage is to compare the offspring and parent population vectors, in the selection stage, to create the parent population for the next iteration. 
Here, the likelihood is used to compare between subunits from the offspring or parent populations, 
%
\begin{equation}
    \mathbf{p}_{*, j} \leftarrow 
    \begin{cases}
        \mathbf{o}_{*, j} & \text{where } \mathcal{L}_{\mathbf{o}_{*, j}} > \mathcal{L}_{\mathbf{p}_{*, j}},\\
        \mathbf{p}_{*, j} & \text{otherwise},
    \end{cases}
\end{equation}
%
where, the $*, j$ subscript notation indicates all objects from the population, $j$.
Following the use of the differential evolution algorithm, typically a more common (gradient-based) approach is used to ensure that the likelihood has been maximised within the space identified by the differential evolution.

@@

The Hessian matrix, $H$ describes the curvature of optimisation criteria, the likelihood, with respect to each of the model parameters, and it is frequently used in the gradient-based optimisation methods. 
This means it is an $i\times i$ matrix, where $i$ is again the number of model parameters, $\theta$, for a two parameter model, this would look like this,
%
\begin{equation}
    H = 
    \begin{bmatrix}
        \dfrac{\partial^2 \mathcal{L}}{\partial \theta_{1}^2} & \dfrac{\partial^2 \mathcal{L}}{\partial \theta_1 \theta_2} \\
        \dfrac{\partial^2 \mathcal{L}}{\partial \theta_2 \theta_1} & \dfrac{\partial^2 \mathcal{L}}{\partial \theta_{2}^2}
    \end{bmatrix}.
\end{equation}
%
If the parameter distribution is Gaussian in shape then the second derivative at the Gaussian maximum will give the variance of that distribution. 
Therefore, when these gradient approaches are used, the Hessian matrix is calculated and the diagonal of that matrix gives the uncertainty in each of the parameters. 
The off-diagonal elements contain information about the covariance between different parameters, however this is rarely leveraged and the covariances are taken to be negligible.

@@

Once an optimised set of model parameters, $\theta$, are obtained which maximise the likelihood, $\mathcal{L}$, some random perturbation is applied, 
%
\begin{equation}
    \Theta = \theta + aR,
\end{equation}
%
where $R$ is some normally distributed number centered on \num{0} with a standard distribution of \num{1} and $a$ is the step size. 
A new $\mathcal{L}$ is found for $\Theta$ and the probability that this transition is will occur is found, 
%
\begin{equation}
    p = \exp{\bigg[\frac{\mathcal{L}(\Theta) - \mathcal{L}(\theta)}{2}\bigg]}. 
\end{equation}
%
This probability is then compared with a uniformly distributed random number from \num{0} to \num{1}, $n$,
%
\begin{equation}
    \theta \leftarrow
    \begin{cases}
        \Theta & \text{where } n < p,\\
        \theta & \text{otherwise}.
    \end{cases}
\end{equation}
%
This process is repeated until some desired number of samples has been obtained. 
It should be noted that it is important to allow the Markov chains to have some ``burn-in'' period, which is not included in the final samples. 
This allows the MCMC algorithm to settle into the search-space.

>>>>

% I know that it is common/precedent to start by looking at the Born approximation, but I wonder if, in paper format, it is just worth pointing out that multiple reflections invalidate Born and something itterative needs to be done? J.C.

@@

When we measure a reflectometry profile, we measure the reflected intensity, and some uncertainty in that measurement, at discrete points in the wavevector, $R(q) \pm \delta R(q)$. 
Using the recursive method discussed above, we can calculate a model reflected intensity at these same $q$ values, $R_m(q)$. 
We then aim to reduce the difference between the measured and modelled reflected intensity through the optimisation (maximisation) of the likelihood, $L$, 
%
\begin{equation}
    L = \exp{\bigg\{-0.5 \sum_{q=q_{\text{min}}}^{q_{\text{max}}} \bigg[\frac{R(q) - R_m(q)}{\delta R(q)}\bigg]^2 + \ln[2\pi \delta R(q)]\bigg\}}.
\end{equation}
%
This parameter described the probability that the model $m$ accurately describes the observed reflectometry data. 
The aim in reflectometry modelling is to obtain a model reflectometry that has the maximum likelihood possible for the given experimental data. 
Figure~\ref{fig:likehood} shows the maximum likelihood model for some experimental data, in addition to another model which does not agree well with the data and therefore has a lower likelihood as a result. 
%
\begin{figure}[t]
    \includegraphics[width=0.49\textwidth]{likelihood}
    \caption{The maximum likelihood fit, the blue data points indicate some experimental data, while the yellow line shows the maximum likelihood model fit to this dataset and the green line show another model (which doesn't maximise the likelihood).}
    \label{fig:likelihood}
\end{figure}
%

@@

The final stage is to compare the offspring and parent population vectors, in the selection stage, to create the parent population for the next iteration. 
Here, the likelihood is used to compare between subunits from the offspring or parent populations, 
%
\begin{equation}
    \mathbf{p}_{*, j} \leftarrow 
    \begin{cases}
        \mathbf{o}_{*, j} & \text{where } L_{\mathbf{o}_{*, j}} > L_{\mathbf{p}_{*, j}},\\
        \mathbf{p}_{*, j} & \text{otherwise},
    \end{cases}
\end{equation}
%
where, the $*, j$ subscript notation indicates all objects from the population, $j$.
Following the use of the differential evolution algorithm, typically a more common (gradient-based) approach is used to ensure that the likelihood has been maximised within the space identified by the differential evolution.

@@

The Hessian matrix, $H$ describes the curvature of optimisation criteria, the likelihood, with respect to each of the model parameters, and it is frequently used in the gradient-based optimisation methods. 
This means it is an $i\times i$ matrix, where $i$ is again the number of model parameters, $\theta$, for a two parameter model, this would look like this,
%
\begin{equation}
    H = 
    \begin{bmatrix}
        \dfrac{\partial^2 L}{\partial \theta_{1}^2} & \dfrac{\partial^2 L}{\partial \theta_1 \theta_2} \\
        \dfrac{\partial^2 L}{\partial \theta_2 \theta_1} & \dfrac{\partial^2 L}{\partial \theta_{2}^2}
    \end{bmatrix}.
\end{equation}
%
If the parameter distribution is Gaussian in shape then the second derivative at the Gaussian maximum will give the variance of that distribution. 
Therefore, when these gradient approaches are used, the Hessian matrix is calculated and the diagonal of that matrix gives the uncertainty in each of the parameters. 
The off-diagonal elements contain information about the covariance between different parameters, however this is rarely leveraged and the covariances are taken to be negligible. 
Additionally, we note that often the Hessian is not calculated in an optimisation so the Jacobian matrix (the first derivative analogue of the Hessian) is used to estimate Hessian matrix.

@@

Once an optimised set of model parameters, $\theta$, are obtained which maximise the likelihood, $L$, some random perturbation is applied, 
%
\begin{equation}
    \Theta = \theta + aR,
\end{equation}
%
where $R$ is some normally distributed number centered on \num{0} with a standard distribution of \num{1} and $a$ is the step size. 
A new $L$ is found for $\Theta$ and the probability that this transition is will occur is found, 
%
\begin{equation}
    p = \exp{\bigg[\frac{L(\Theta) - L(\theta)}{2}\bigg]}. 
\end{equation}
%
This probability is then compared with a uniformly distributed random number from \num{0} to \num{1}, $n$,
%
\begin{equation}
    \theta \leftarrow
    \begin{cases}
        \Theta & \text{where } n < p,\\
        \theta & \text{otherwise}.
    \end{cases}
\end{equation}
%
This process is repeated until some desired number of samples has been obtained. 
It should be noted that it is important to allow the Markov chains to have some ``burn-in'' period, which is not included in the final samples. 
This allows the MCMC algorithm to settle into the search-space.
